Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-19 index.qmd
UPdates

@@

In order to use CPH to model time to event survival data, the data must contain a value for time, a binary field indicating the event being modeled, an id field, and one or multiple covariate fields. The time value is typically formatted as a time series for each individual in the dataset. Individuals will have a time value from the beginning of the observation period, to an arbitrary time that may or may not be the time the event being modeled for occurs. Data is considered censored when the individual is not given the opportunity to experience the event in the observation period and is handled differently by the model. The covariate fields in the dataset can be categorical or numerical. [@nahhas2025].

@@

#### Generating Model
To generate the CPH model with properly formatted survival data, a survival object must first be created using the R package `survival`. The survival object is created by running `Surv(time, event)` where `time` is the time field and `event` is the event field.

@@

The proportional hazards assumption has been proven to be critical to
evaluate prior to using CPH when modeling survival data. The
proportional hazards assumption states that the ratio of hazard rates
for any two individuals must be constant over time regardless of the
point in time of the measurement. This means the effect
of the covariates must be constant over time. [@Bustan2018]

@@

This assumption describes how if an event is censored it should not be considered when
the model is being created. Censored data is difficult to accommodate when
Cox proportional hazards models are used in other industries, such as
the medical field, because observation times for the individuals
involved in studies may have decades before an event is observed and the event may not have the opportunity to occur during the study period. In
this scenario, it would not be appropriate to consider this individual
when modeling survival.

@@

Survival times of observed individuals should be independent of each
other. In an example where survival of plane engines were being evaluated, if each plane has two engines and the failure of one engine caused the accelerated failure of another due to increased workload, it would produce biased conclusions and violate the independence assumption.

@@

#### Forest Plot
Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio. The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis. The plot contains whiskers demonstrating a 95% confidence interval for the effect. A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.

@@

#### Hazard Ratio

CPH models focus on providing a relative risk of the event happening based on the covariates and value for time. The hazards ratio is used to evaluate this level of risk and is calculated by finding the ratio between two individuals where one is the baseline value of a covariate at time $t$, and the other is the value of a covariate with a one unit increase. The hazards ratio is represented by the formula below.

$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] = \text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$

-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values.

A hazards ratio can be interpreted as:
$h(t|\mathbf{Z})$ < 1 indicates a decreased baseline risk as compared to the baseline.
$h(t|\mathbf{Z})$ > 1 indicates an increased baseline risk as compared to the baseline.

Hazard ratios can be interpreted to give an estimation of the event occurring. For example, a hazard ratio of 1.3 signifies that for a one unit increase in the covariate, a 30% increase in risk of the event occurring is expected. Hazard ratios can also be used to calculate hazard rates, which have theiur own applications and interpretations when modeling survival data. [@klein2005]

#### Hazard Rate

The hazard rate is defined as:

$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}({\sum_{k=1}^{p} \beta_kZ_k}) $$

-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
    with risk vector $\mathbf{Z}$

-   $\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)$ is the hazard ratio as explained above.

-   $h_0(t)$ is a baseline hazard rate indicating the hazard rate when all covariates are zero.

The resulting hazard rate is a measure of the instantaneous risk of the event happening at a specific point in time. Similar to the hazard ratio, hazard rates are interpreted as:

$h(t|\mathbf{Z})$ < 1 indicates a lower risk of the event occurring at the specific point in time.
$h(t|\mathbf{Z})$ > 1 indicates a higher risk of the event occurring at the specific point in time.

Hazard rates can be used as an additional measure of risk, as well as to calculate the cumulative hazard for an individual. [@klein2005]

The hazard rate can be calculated in R by using `stats::predict()` after creating a CPH model. This function will be used to create a column showing the hazard rate for each row in the testing data.

#### Cumulative Hazard

Cumulative hazard represents the total accumulated risk of an individual for the duration of the observation period at time $t$ and is defined as:

$H(t) = \int_0^t h(t) dt$
where,

-   $H(t)$ is the cumulative hazard at time $t$

-   $h(t)$ is the hazard rate at time t

As defined by the formula, cumulative hazard is a summation of the hazard rates for an individual from time 0 to time $t$. A higher cumulative hazard indicates a higher risk of event. Cumulative hazard is used as an additional measure for risk, but it is mainly used to calculate survival probability to plot against time on survival curves. [@klein2005]

The cumulative hazard will be calculated in R by grouping the data by the id field and finding the summation of the hazard rates for the individual through the duration of the observation period. A column will be added showing the cumulative hazard for each individual at each point in time.

#### Calculating Survival Probability

$S(t) = exp(-H(t))$

-   $S(t)$ is the probability of survival for an individual at time $t$
-   $H(t)$ is the cumulative hazard for an individual at time $t$

The resulting survival probability will be between 0 and 1 with a value close to one indicating a high probability of the individual not experiencing the event. Conversely, a value close to 0 indicates a higher probability of the individual experiencing the event. A survival probability of .05 indicates that only 5% of individuals under the same conditions would survive. Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.[@klein2005]

@@

```{r}
plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```
survival curve

>>>>

### Mathematical Formulas

There are two functions that are the backbone for CPH: the survival
function and the hazard function. The survival function when $X$ is a
continuous random variable is

$$
S(x) = 1 - F(x) = 1 - Pr(X > x) = \int_x^\infty f(x) dx
$$ {#eq-survival}

where $S(x)$ is the probability that an individual has survived past
time $x$ [@klein2005]. undefined

The hazard function, which is also known as the hazard rate function,
can be defined as

$$
 h(x) = \lim_{\Delta x \rightarrow 0} \frac{P[x\leq X < x + \Delta x | X \geq x]}{\Delta x}
$$ {#eq-hazardrate}

The function is used to describe the changes in [@klein2005].

Since CPH is a proportional hazard model, the hazard function is adapted
to deal with the relationship between time to event and the explanatory
variables [@klein2005]. It can be defined as

$$
h(t|\mathbf{Z} ) = h_0(t)\text{exp}(\sum_{k=1}^p \beta_kZ_k)
$$ {#eq-coxhazard}

-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an subject with
    risk vector $\mathbf{Z}$
-   $h_0(t)$ is a baseline hazard rate when all covariates are $0$
-   $\text{exp}(\sum_{k=1}^p \beta_k Z_k)$ is a semi-parametric model
-   $\beta_k$ is a parameter vector
-   $Z_k$ is the risk vector [@klein2005]

CPH is considered a proportional hazards model due to the hazard ratio
between two individuals being a constant value. The ratio of the hazard
function for an individual with $\mathbf{Z}$ covariates where
$Z_k = z+1$ and for an individual with $\mathbf{Z}^*$ covariates where
$Z_k = z$ with all other predictors are fixed [@nahhas2025] is defined
as

$$
\frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}[\sum_{k=1}^p \beta_k Z_k]}{h_0(t)\text{exp}[\sum_{k=1}^p \beta_k Z^*_k]} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k - Z_k^*)]
$$ {#eq-hazardratio}[@klein2005]. The result of the hazard ratio (HR) is
a constant that implies an association between a continuous predictor
and the outcome. The interpretation of the HR is

-   if HR $> 1$ implies a positive association such that
    $100\% \times (HR - 1)$ is the \\% of greater risk for $Z = z+1$
    when compared to $Z = z$.
-   if HR $< 1$ implies a negatie association such that
    $100\% \times (HR -1)$ is the \\% of decreased risk for $Z = z+1$
    when compared to $Z= z$.
-   if HR $=1$ implies no association. [@nahhas2025]

An additional useful equations is the cumulative hazard function. The
cumulative hazard function can be defined as

$$
 H(x) = \int_0^x h(u) du 
$$ {#eq-cumulativehr}

where $h(u)$ is the hazard rate function [@klein2005]. This function
provides a better understanding of the survival function for graphing
purposes.

@@

In order to use CPH to model time to event survival data, the data must
contain a value for time, a binary field indicating the event being
modeled, an id field, and one or multiple covariate fields. The time
value is typically formatted as a time series for each subject in the
dataset. Subjects will have a time value from the beginning of the
observation period, to an arbitrary time that may or may not be the time
the event being modeled for occurs. Data is considered censored when the
subject is not given the opportunity to experience the event in the
observation period and is handled differently by the model. The
covariate fields in the dataset can be categorical or numerical
[@nahhas2025].

@@

#### Generating Model

To generate the CPH model with properly formatted survival data, a
survival object must first be created using the R package `survival`.
The survival object is created by running `Surv(time, event)` where
`time` is the time field and `event` is the event field.

@@

The proportional hazards assumption has been proven to be critical to
evaluate prior to using CPH when modeling survival data. The
proportional hazards assumption states that the ratio of hazard rates
for any two subjects must be constant over time regardless of the point
in time of the measurement. This means the effect of the covariates must
be constant over time. [@Bustan2018]

@@

This assumption describes how if an event is censored it should not be
considered when the model is being created. Censored data is difficult
to accommodate when Cox proportional hazards models are used in other
industries, such as the medical field, because observation times for the
subjects involved in studies may have decades before an event is
observed and the event may not have the opportunity to occur during the
study period. In this scenario, it would not be appropriate to consider
this subject when modeling survival.

@@

Survival times of observed subjects should be independent of each other.
In an example where survival of plane engines were being evaluated, if
each plane has two engines and the failure of one engine caused the
accelerated failure of another due to increased workload, it would
produce biased conclusions and violate the independence assumption.

@@

#### Forest Plot

Forest plots can be used to visualize the effects of each covariate in
the model on the hazard ratio. The resulting plot contains each
covariate on the y axis, with the estimated effect on the hazard ratio
on the the x axis. The plot contains whiskers demonstrating a 95%
confidence interval for the effect. A positive effect indicates a
positive correlation with the hazard with a higher number signifying a
more severe effect on the hazard ratio.

@@

@@

```{r}
plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

survival curve
